mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-11 Draft3.Rmd
table

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R\ functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R\ code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic representation, this new
  feature is a futher step towards reproducible science. User-defined hooks
  allow to extend the package by mapping new variable names and functions to
  new MathML and LaTeX entities. We provide a few working examples that
  illustrate the use of the package in a little scientific report and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
    tables: yes
bibliography: RJreferences.bib
---

@@

```{r custom-operators, echo=FALSE}
op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
  integerdivision=quote(x %/% y))
m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
t <- cbind(Operator=as.character(op), Output=m)
knitr::kable(t, caption = "Custom operators in mathml")
```

@@

More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
translated to the form `varphi(X; M, S)`.

```{r, results='asis'}
hook(call("dnorm", expression(X), expression(M), expression(S)), 
     call("fn", quote(varphi),
       call(";", list(expression(X)), list(expression(M), expression(S)))))
term <- quote(dnorm(x, mu, sigma^2L))
mathout(term)
```

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R\ functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R\ code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic representation, this new
  feature is a futher step towards reproducible science. User-defined hooks
  allow to extend the package by mapping new variable names and functions to
  new MathML and LaTeX entities. We provide a few working examples that
  illustrate the use of the package in a little scientific report and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: \usepackage{longtable}
---

@@

```{r custom-operators, echo=FALSE}
op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y))
m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

op <- as.character(op)
if(knitr::is_latex_output())
  op <- sapply(op, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op <- sapply(op, FUN=knitr:::escape_html)

t <- cbind(Operator=op, Output=m)
knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
```

@@

More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
translated to the form `phi(X; M, S)` (instead of the default `varphi`).

```{r, results='asis'}
hook(call("dnorm", expression(X), expression(M), expression(S)), 
     call("fn", quote(phi),
       call(";", list(expression(X)), list(expression(M), expression(S)))))
term <- quote(dnorm(x, mu, sigma^2L))
mathout(term)
```
